A weakly structured stem for human origins in Africa

### 2022-01-09 paper.tex
edits to results

\begin{figure}[ht!]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/fig1.pdf}}
\caption{\textbf{Geography and genetic diversity within Africa}.
    \aprcomment{In the models presented below, we focus on the Nama, 
    Mende (MSL), Gumuz, and Oromo and Amhara.}
}
\label{fig:1}
\end{center}
\end{figure}

@@

We began with a model of geographic expansion from a single ancestral source
followed by migration between populations
(Figure~\ref{fig:supp-possible-models}A), without allowing for contribution
from an archaic African lineage or population structure prior to the expansion.
As expected \citep{Ragsdale2019-nt}, this first model was a poor fit to the
data qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and
quantitatively (log-likelihood ($LL$) $\approx -189,400$,
Table~\ref{tab:supp-single-origin}). We next explored a suite of models in
which population structure predates the differentiation of contemporary groups,
including models allowing for ancestral reticulation
(Figure~\ref{fig:supp-possible-models}B), archaic admixture
(Figure~\ref{fig:supp-possible-models}C), and African multi-regionalism
(Figure~\ref{fig:supp-possible-models}D).

Regardless of the model choice for early epochs, inference of human
demographic history in the last 150kya was remarkably robust. The earliest divergence
among contemporary sampled human populations differentiates the southern
African Nama from other African groups between 110--135kya, with low to moderate
levels of subsequent gene flow (Table\ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140kya. We conclude that, regardless of model
specification, geographic patterns of \emph{Homo sapiens} population structure
date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

@@

\subsection*{Deep population structure but not archaic admixture within Africa}

@@

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (kya)} & \textbf{per generation} & \textbf{duration (ky)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        LL= -115,500 & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        LL= -102,600 & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
        & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
        & g, & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}
        Labeled migration rates correspond to symmetric continuous migration
        bands shown in Figure~\ref{fig:2}. Both the continuous-migration and
        merger models \ldots
        \aprcomment{also define *}
    }
    \label{tab:migration-rates}
\end{table}

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration (e.g. $LL \approx -102,600$ vs.
$-107,700$ in the merger scenario and $LL \approx -115,500$ vs. $-128,000$ with
continuous migration). With continuous migration between stems, population
structure extends back to 1.1--1.4Mya (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this roughly compares
to inferred migration rates between connected contemporary populations over the
past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or, at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic admixture models (Figure~\ref{fig:supp-proposed-models}C) as
the early branches remain closely related (Figure 3). \aprcomment{this last
phrase may need to be updated, with regard to Fig \ref{fig:supp-FST-panels}}

Bootstrap-based confidence intervals are relatively narrow within each
specified model
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. As expected, model assumptions
have a larger impact on parameter estimates. To convey model uncertainty, we
highlight features of two models with high likelihoods. These differ primarily
in whether early stems are assumed to give rise to modern populations in
multiple discrete merger events or in continuous migrations
(Figure~\ref{fig:2}). Additional models are discussed in the Supporting
Material.

The “Multiple Merger” model has better log-likelihood than the “Continuous
Migration” model (-102,600 vs. -115,500, respectively). The two models differ
in the timing of divergence and the relative Ne of the two stems
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

Under the continuous-migration model, the two stems give rise to contemporary
populations in western, southern and eastern Africa but with differing
proportions of ancestry derived from “Stem 1” and “Stem 2” (Figure 2). For
example, the Nama and Gumuz have lower levels of migration with Stem 2
($m=5.8\times10^{-5}$ and $3.1\times10^{-5}$, respectively), relative to the
Mende. The Mende continue to exchange migrants with Stem 2
($m=1.6\times10^{-4}$) long after their separation 60kya; this is a similar
level of gene flow as between the Mende and Eastern Africans (Table 1).

In contrast, under the Multiple Merger model, we allowed for stem populations
to have large instantaneous contributions, in different proportions, to form
the different contemporary groups. However, the ghost populations do not neatly
line up one-to-one with southern, western and eastern populations. We observe a
deep bottleneck in Stem 1 down to $N_e=117$. This represents the lower bound
allowed in our optimization (i.e. 1\% of the $N_{ancestral}$) and the size of
this bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period
of exchange with Stem 2, Stem 1 then fractures into “Stem 1E” and “Stem 1S”
479kya. The timing of this divergence was also poorly constrained ($\sigma_T=
166$kya). These populations evolve independently until approximately 119kya
when Stem 1S and Stem 2 merge to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 merge to
form the ancestors of the Western Africans and Eastern Africans (and thus also
all individuals who later disperse during the Out of Africa event). The
ancestral Western / Eastern African group derives equal ancestry from the two
stems (50\% from Stem 1E and Stem 2, Table 1). Finally, the Mende receive a
large pulse of gene flow from Stem 2, replacing 18\% of their population 25kya.
The later Stem 2 contribution to the Western African Mende resulted in better
model fits (LL). This may indicate that an ancestral Stem 2 population
occupied Western or Central Africa, broadly speaking. The differing proportions
in the Nama and Eastern Africans may also indicate geographic separation of
Stem 1S in southern Africa and Stem 1E in eastern Africa.

@@

Previous studies have found support for archaic admixture in Africa using
two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
(cSFS) \citep{Durvasula2020-td}, and reconstruction of gene genealogies
\citep{Speidel2019-nj}. \aprcomment{Will need to check how strongly the Relate
paper claims that this supports archaic admixture vs simply deep structure. It
wasn't a model fit, in any case.} However, none of these studies considered a
weakly structured stem. We validated our Continuous Migration and Multiple
Merger models with additional independent approaches. We find that the two
models describe the cSFS accurately \aprcomment{(Figure SX)}, even though this
statistic was not used in the fit.  Both models outperform archaic models fit
directly to the cSFS (for example, compare with Figure~1 in
\citet{Durvasula2020-td}).

Similarly, we generated estimates of effective population size as inferred from
coalescence rates with Relate. Changes in Ne over time from prior studies
strongly support a bottleneck in Out of Africa populations, but African
populations also dip in effective size around 200kya, although this decrease is
more modest than the OOA event \citep{Mallick2016-lx}. One possible explanation
is that ancestral population structure during the Middle Stone Age inflated Ne
and subsequent gene flow among ancestral Africans resulted in a lower Ne after
100-200kya \citep{Mazet2016-wn}\aprcomment{Is this a good ref here?}. To our
knowledge, explicit comparisons of this hypothesis with SMC-style approaches
have not been published. We simulated genomic data under 4 models
\aprcomment{(Figure SX)}. All models, including the Single Origin model,
generate multi-modal curves indicating that the ancestral increase in Ne
between 100kya-1Mya is a general feature of the coalescent rate in humans.
Simulated Relate Ne curves under the CM and MM are distinguished by having 2
and 3 modes, respectively. The Continuous Migration model had the better
qualitative fit with the data, but the ancestral mode was shifted further back
in time relative to the data. All simulated models underestimated the level of
population growth during the Holocene, indicating additional free parameters
are necessary to refine these growth rates. \aprcomment{Discussion on this
paragraph moved to issues (issue \#1)}

Relative cross-coalescence rates (rCCR) have recently been used to estimate
divergence between a pair of populations, as measured by the rate of
coalescence between two groups divided by the mean within population
coalescence. Simulations of rCCR accuracy, however, focus on a ‘clean split’
between populations whereby groups diverge without subsequent gene flow.
Published estimates of the earliest human divergences with rCCR, which range
from 150kya-100kya \citep{Bergstrom2021-iw}, may be significantly biased when
compared to more complex models with gene flow as inferred here. We find that
midpoint estimates of rCCR are poor estimates for population divergence, often
underestimating divergence time by 50\% or greater (e.g. Mende vs. Gumuz ~15kya
compared to a true divergence 60kya, \aprcomment{Supp Figure SX}). We suggest
that rCCR analyses which do not fit multiple parameters including gene flow
should be interpreted with caution.

>>>>

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/fig1.pdf}
    \caption{
        \textbf{Geography and genetic diversity within Africa.}
        The combined 1000 Genomes and AGRP datasets incude populations from
        West, East, and South Africa. Both PCA (B, C) and \texttt{ADMIXTURE}
        (D) illustrate the well-characterized high levels of genetic diversity
        and signatures of recent migration and gene flow within Africa.
        We built parameterized models using the newly-sequenced Nama, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderhtal individual (bold fonts).
        \aprcomment{work on this caption}
    }
    \label{fig:1}
\end{figure}

@@

We began with a model of geographic expansion from a single ancestral,
unstructured source followed by migration between populations, without allowing
for contribution from an archaic African lineage or population structure prior
to the expansion (Figure~\ref{fig:supp-possible-models}A). As expected
\citep{Ragsdale2019-nt}, this first model was a poor fit to the data
qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and quantitatively
(log-likelihood ($LL$) $\approx -189,400$, Table~\ref{tab:supp-single-origin}).
We next explored a suite of models in which population structure predates the
differentiation of contemporary groups, including models allowing for ancestral
reticulation (Figure~\ref{fig:supp-possible-models}B), archaic admixture
(Figure~\ref{fig:supp-possible-models}C), and African multi-regionalism
(Figure~\ref{fig:supp-possible-models}D).

Regardless of the model choice for early epochs, inference of human demographic
history in the last 150kya was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135kya, with low to moderate levels of
subsequent gene flow (Table\ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140kya. We conclude that, regardless of model
specification, geographic patterns of \emph{Homo sapiens} population structure
date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

@@

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (kya)} & \textbf{per generation} & \textbf{duration (ky)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        LL= -115,500 & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        LL= -102,600 & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
        & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
        & g, & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}
        Labeled migration rates correspond to symmetric continuous migration
        bands shown in Figure~\ref{fig:2}. Both the continuous-migration and
        merger models inferred a relatively deep split of human stem branches,
        though they were connected by ongoing migration that maintained their
        genetic similarity.
        *: The ancestors of the Nama diverged from other African groups
        $\sim$120--135kya, which later branch into West and East African
        groups 60kya. Migration rates and durations are shown for 1) East Africans
        and their ancestors, and 2) the Mende, respectively.
    }
    \label{tab:migration-rates}
\end{table}

\subsection*{Deep but connected population structure within Africa}

@@

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration ($LL \approx -102,600$ vs.
$-107,700$ in the merger scenario and $LL \approx -115,500$ vs. $-128,000$ with
continuous migration). With continuous migration between stems, population
structure extends back to 1.1--1.4Mya (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this roughly compares
to inferred migration rates between connected contemporary populations over the
past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic admixture models (Figure~\ref{fig:supp-possible-models}C) as
the early branches remain closely related (Figure 3).

Bootstrap-based confidence intervals are relatively narrow within each
specified model
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. As expected, model assumptions
have a larger impact on parameter estimates. To convey model uncertainty we
highlight features of two inferred models with high likelihoods, the
“multiple-merger” and the the “continuous-migration” model that both allow
migration between stem branches. These differ primarily in the timing of the
early divergence of stem populations and their relative $N_e$
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

Under the continuous-migration model, one of the two stems diverges into
lineages leading to contemporary populations in western, southern and eastern
Africa, and the other (Stem 2) contributes variable ancestry to those
populations. This migration from Stem 2 is highest with the Mende
($m=1.6\times10^{-4}$) compared to the Nama and East African populations
($m=5.8\times10^{-5}$ and $3.1\times10^{-5}$, respectively), with migration
allowed to occur until the relatively recently demographic events 5kya. A
sampled lineage from the Nama, Mende, and Gumuz have probabilities of being in
Stem 2 at the time of Stem 1 expansion (135ka) of approximately $0.145$, $0.2$,
and $0.13$, respectively, though these probabilities change over time,
precluding the notion of a fixed admixture proportion.

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
%However, the ghost populations do not neatly line up one-to-one with southern,
%western and eastern populations.
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into “Stem 1E” and “Stem 1S”
479kya. The timing of this divergence was also poorly constrained ($\sigma_T=
166$kya). These populations evolve independently until approximately 119kya
when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the Western Africans and
Eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25kya. The later Stem
2 contribution to the Western African Mende resulted in better model fits (LL).
This may indicate that an ancestral Stem 2 population occupied Western or
Central Africa, broadly speaking. The differing proportions in the Nama and
Eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

Previous studies have found support for archaic admixture in Africa using
two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
(cSFS) \citep{Durvasula2020-td}, and reconstruction of gene genealogies
\citep{Speidel2019-nj}. However, none of these studies considered a weakly
structured stem. We validated our inferred models with additional independent
approaches. We find that the observed cSFS (conditioned on the derived allele
being carried in the Neanderthal sample) is very well-described by the merger
model (Figures~\ref{fig:4}A-C and
\ref{fig:supp-csfs-single-origin}--\ref{fig:supp-csfs-merger-with-stem-migration}),
even though this statistic was not used in the fit. Both models outperform
archaic models fit directly to the cSFS (for example, compare with Figure~1 in
\citet{Durvasula2020-td}).

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time, which are commonly interpreted as effective
population size histories. Trajectories of $N_e$ from prior studies strongly
support a bottleneck in Out of Africa populations. African populations also dip
in effective size around 100-200kya, although this decrease is more modest than
the OOA event \citep{Mallick2016-lx}. One possible explanation is that
ancestral population structure during the Middle Stone Age inflated $N_e$ and
subsequent gene flow among ancestral Africans resulted in a more recent
decreased $N_e$ \citep{Mazet2016-wn}.
%To our knowledge, explicit comparisons of this hypothesis with SMC-style
%approaches have not been published.
We simulated genomic data under four inferred demographic models to compare
with observed coalescence rate histories. All models, including the
single-origin model, generate multi-modal curves indicating that the ancestral
increase in $N_e$ between 100kya-1Mya is a general feature of the coalescent
rate in humans, although none of the models perfectly match the exact timing of
the early fluctuations in $N_e$ observed in the data.
%Simulated Relate Ne curves under the CM and MM are distinguished by having 2
%and 3 modes, respectively.  The Continuous Migration model had the better
%qualitative fit with the data, but the ancestral mode was shifted further back
%in time relative to the data.
Each simulated model also underestimated the level of population growth during
the Holocene, indicating additional free parameters are necessary to refine
these growth rates.

Relative cross-coalescence rates (rCCR) have recently been used to estimate
divergences between a pair of populations, as measured by the rate of
coalescence between two groups divided by the mean within population
coalescence. Simulations of rCCR accuracy, however, focus on a ‘clean split’
between populations whereby groups diverge without subsequent gene flow.
Published estimates of the earliest human divergences with rCCR, which range
from 150kya-100kya \citep{Bergstrom2021-iw}, may be significantly biased when
compared to more complex models with gene flow as inferred here. We find that
midpoint estimates of rCCR are poor estimates for population divergence, often
underestimating divergence time by 50\% or greater (e.g., Mende vs. Gumuz
$\sim$15kya compared to a true divergence of 60kya), and recent migration can
lead to the misordering of divergence events (Figure~\ref{fig:4}E). We suggest
that rCCR analyses which do not fit multiple parameters including gene flow
should be interpreted with caution.
